A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-20 tex/codatadp.tex
Finish first draft of lemma.

\begin{lemma}
\label{lem:mu-norm-arg}
If $\vec u$ are in normal form and $t = \nf{\MU x.\; \const{C}( \vec u )} = \MU x.\; \const{C}( \vec w )$,
then $\vec u \vsim t\interp{\vec w}$.
\end{lemma}
\begin{proof}
%Since $u_k$ is normal, $w_k$ is the result of replacing all subterms $v$ of $u_k$ such that $t\interp{v} \aequiv t$ with $x$.
\rem{TODO}
\end{proof}

@@

\begin{lemma}
\label{lem:model-completion}
If $\Val$ is constructed for a saturated set $\Ec$
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, the following holds:
\begin{enumerate}
\item[\rm 1.] $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$, where $\ec{x}$ has type $\tau$.
\item[\rm 2.]
All subterms $t$ of $\Val^\ast \ec{y}$ with $\muvar(t) = {\Varec{x}}$ are such that
$\Val^\ast \ec{y} \interp{ t } = \Val^\ast \ec{x}$. 
%If $\const{C}( t_1, \ldots, t_n ) \in \ec{x}$,
%then all subterms $\MU v_{\ec{x}}.\; u$ occurring in $\Val^\ast \ec{y}$ 
%are of the form $\MU v_{\ec{x}}.\; \const{C}( w_1, \ldots, w_n )$,
%where 
%and $\Val^\ast \ec{y} \interp{ w_i } = \Val^\ast \ec{t_i}$ for $i = 1, \ldots, n$.
%If $\Val^\ast \ec{x}$ contains a subterm of the form $\MU v_{\ec{x}}.\; \const{C}( w_1, \ldots, w_n )$,
%and $\muvar( w_i ) = v_{\ec{t_i}}$ for $i = 1, \ldots, n$,
%\item
%$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
%for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)={\Varec{x}}$,
\item[\rm 3.]
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable ${\Varec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else ${\Varec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be $1$.
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_x(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to ${\Varec{x}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction,
since there are only a finite number of closed terms in $\mathcal{T}^\M_x(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_x(\Val^\ast)$ to ${\Varec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form
$\{ y \mapsto t \}$, where $t$ is acyclic when the type of $y$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$.

We first show that 2 and 3 hold initially for $\Val$.
For all equivalence classes $\ec{z}$,
each pair of constructor terms $\const{C}_j( \vec t )$ and $\const{C}_{j'}( \vec u )$ in $\ec{z}$
are such that $j = j'$ since \rn{Clash} does not apply to $\Ec$,
and are such that $\ec{ \vec t } = \ec{ \vec u }$ since \rn{Inject} does not apply to $\Ec$.
Thus, 
$\Val$ was constructed by applying a sequence of substitutions
where all substitutions for variables ${\Varec{z}}$ 
were uniquely of the form $\{ {\Varec{z}} \mapsto \const{C}_j( {\Varec{t_1}}, \ldots, {\Varec{t_n}} ) \}$
when $\ec{z}$ contains a constructor $\const{C}_j( t_1, \ldots, t_n )$.
Say $\Val \ec{y}$ has a subterm $t$ where $\muvar(t) = {\Varec{x}}$.
Both $\Val \ec{x}$ and the subterm $t$ of $\Val \ec{y}$ were
constructed by applying a sequence of substitutions of the form mentioned above to ${\Varec{x}}$.
Moreover, free variables ${\Varec{z}}$ in $\MU {\Varec{x}}.\; t$ that are bound in $\Val \ec{y}$ are interpreted in 
the expansion of $\Val \ec{y} \interp{ \MU {\Varec{x}}.\; t }$ as a term
constructed by a sequence of substitutions of the form mentioned above to ${\Varec{z}}$. \rem{more detail}
Thus, we have
$\Val \ec{y} \interp{ \MU {\Varec{x}}.\; t } = \Val \ec{x} \interp{ \Val \ec{x} } = \Val \ec{x}$, and thus 2 holds for $\Val$.
Statement 3 holds for $\Val$ since \rn{Unique} does not apply.

We now show that $\Val = \nf{(\Val)}$.
Assume by contradition $\Val \ec{x} \neq \nf{(\Val \ec{x})}$ for some $\Val \ec{x}$ of minimal size.
We have that $\Val \ec{x}$ is of the form $\MU {\Varec{x}}.\; \const{C}( t_1, \ldots, t_n )$.
Due to our construction of $\Val$, 
we know $\ec{x}$ contains a constructor $\const{C}( z_1, \ldots, z_n )$ and $\muvar( t_i ) = z_i$ for $i = 1, \ldots, n$.
Since $\Val \ec{x}$ is a minimal, it
contain a subterm of the form $\MU {\Varec{y}}.\; \const{C}( u_1, \ldots, u_n )$
where $\Val \ec{x} \interp{ t_i } \aequiv \Val \ec{x} \interp{ u_i }$ for $i = 1, \ldots, n$.
Due to our construction of $\Val$, $\ec{y}$ contains a constructor $\const{C}( w_1, \ldots, w_n )$ and $\muvar( u_i ) = w_i$ for $i = 1, \ldots, n$.
Since \rn{Cong} does not apply to $\Ec$
we have $\ec{w_j}$ and $\ec{z_j}$ are distinct for some $j$.
By statement 2, $\Val \ec{x} \interp{ t_j } = \Val \ec{z_j}$
and $\Val \ec{x} \interp{ u_j } = \Val \ec{w_j}$,
which are not $\alpha$-equivalent by statement 3, contradicting the fact that $\MU {\Varec{y}}.\; \const{C}( u_1, \ldots, u_n )$ is a bisimilar subterm of $\Val \ec{x}$.
Thus, $\Val = \nf{(\Val)}$, and statements 2 and 3 hold for $\nf{(\Val)}$.

We now show that if statements 2 and 3 hold for some $\Val_1$,
then 2 and 3 also hold for $\nf{(\Val_1 \sigma)}$,
where $\sigma$ is a substitution of the form $\{ {\Varec{x}} \mapsto \mu {\Varec{x}}. t \}$
and $\mu {\Varec{x}}. t$ is not $\alpha$-equivalent to a term in $\mathcal{T}^\M_{\Varec{x}}(\Val_1)$.
To show 2,
by assumption of 2 on $\Val_1$, we have $\Val_1 \ec{y} \interp{ t } = \Val_1 \ec{x}$ for
all subterms $t$ of $\Val_1 \ec{y}$ where $\muvar(t) = {\Varec{x}}$.
Thus, $\Val_1 \sigma \ec{y} \interp{ t } = \Val_1 \sigma \ec{x}$ and
$\nf{(\Val_1 \sigma)} \ec{y} \interp{ t } = \nf{(\Val_1 \sigma)} \ec{x}$.
To show 3, 
consider two distinct equivalence classes $\ec{x}$ and $\ec{y}$,
and assume by contradiction that $\Val_1 \sigma \ec{x} \aequiv \Val_1 \sigma \ec{y}$.
Due to statement 3 for $\Val_1$,
$\ec{y}$ and $\ec{z}$ must have (minimal) subterms 
where $t_1$ occurs in $\Val_1 \ec{y}$ the same position as $t_2$ occurs in $\Val_1 \ec{z}$,
and $t_1 \not\aequiv t_2$.
If $t_1$ and $t_2$ are distinct free variables,
then $t_1 \sigma \not \aequiv t_2 \sigma$.
If $t_1$ is of the form $\mu {\Varec{w_1}}. \const{C}( \vec t )$ 
and $t_2$ is of the form $\mu {\Varec{w_2}}. \const{D}( \vec u )$,
then $t_1 \sigma \not \aequiv t_2 \sigma$.
Since $t_1$ and $t_2$ are minimal,
$t_1$ must be of the form $\mu {\Varec{z}}. \const{C}( \vec t )$,
and $t_2$ must be a free variable.
If $t_2$ is not ${\Varec{x}}$,
then $t_1 \sigma \not \aequiv t_2 \sigma$.
Thus, $t_2$ is ${\Varec{x}}$.
\rem{Must show positions do not collapse.}
\qed
\end{proof}

>>>>

\begin{lemma}
\label{lem:mu-norm-arg}
If $\vec u$ are in normal form and $t = \nf{\MU x.\; \const{C}( \vec u )} = \MU x.\; \const{C}( \vec w )$,
then $\vec u \vsim t\interp{\vec w}$.
\end{lemma}
\begin{proof}
\rem{TODO}
\end{proof}

\begin{lemma}
\label{lem:mu-norm-interp}
If $t$ is normal and has subterm $u$, then $t \interp{u}$ is normal.
\end{lemma}
\begin{proof}
\rem{TODO}
\end{proof}

@@

\begin{lemma}
\label{lem:model-completion}
If $\Val$ is constructed for a saturated set $\Ec$
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, the following holds:
\begin{enumerate}
\item[\rm 1.] $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$, where $\ec{x}$ has type $\tau$.
\item[\rm 2.]
All subterms $t$ of $\Val^\ast \ec{y}$ with $\muvar(t) = {\Varec{x}}$ are such that
$\Val^\ast \ec{y} \interp{ t } = \Val^\ast \ec{x}$. 
%If $\const{C}( t_1, \ldots, t_n ) \in \ec{x}$,
%then all subterms $\MU v_{\ec{x}}.\; u$ occurring in $\Val^\ast \ec{y}$ 
%are of the form $\MU v_{\ec{x}}.\; \const{C}( w_1, \ldots, w_n )$,
%where 
%and $\Val^\ast \ec{y} \interp{ w_i } = \Val^\ast \ec{t_i}$ for $i = 1, \ldots, n$.
%If $\Val^\ast \ec{x}$ contains a subterm of the form $\MU v_{\ec{x}}.\; \const{C}( w_1, \ldots, w_n )$,
%and $\muvar( w_i ) = v_{\ec{t_i}}$ for $i = 1, \ldots, n$,
%\item
%$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
%for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)={\Varec{x}}$,
\item[\rm 3.]
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable ${\Varec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else ${\Varec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be $1$.
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_{\Varec{y}}(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to ${\Varec{y}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction,
since there are only a finite number of closed terms in $\mathcal{T}^\M_{\Varec{y}}(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_{\Varec{y}}(\Val^\ast)$ to ${\Varec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form
$\{ {\Varec{y}} \mapsto t \}$, where $t$ is acyclic when the type of ${\Varec{y}}$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$.

We first show that 2 and 3 hold initially for $\Val$.
For all equivalence classes $\ec{z}$,
each pair of constructor terms $\const{C}_j( \vec t )$ and $\const{C}_{j'}( \vec u )$ in $\ec{z}$
are such that $j = j'$ since \rn{Clash} does not apply to $\Ec$,
and are such that $\ec{ \vec t } = \ec{ \vec u }$ since \rn{Inject} does not apply to $\Ec$.
Thus, 
$\Val$ was constructed by applying a sequence of substitutions
where all substitutions for variables ${\Varec{z}}$ 
were uniquely of the form $\{ {\Varec{z}} \mapsto \const{C}_j( {\Varec{t_1}}, \ldots, {\Varec{t_n}} ) \}$
when $\ec{z}$ contains a constructor $\const{C}_j( t_1, \ldots, t_n )$.
Say $\Val \ec{y}$ has a subterm $t$ where $\muvar(t) = {\Varec{x}}$.
Both $\Val \ec{x}$ and the subterm $t$ of $\Val \ec{y}$ were
constructed by applying a sequence of substitutions of the form mentioned above to ${\Varec{x}}$.
Moreover, free variables ${\Varec{z}}$ in $t$ that are bound in $\Val \ec{y}$ are interpreted in 
the expansion of $\Val \ec{y} \interp{ t }$ as a term
constructed by a sequence of substitutions of the form mentioned above to ${\Varec{z}}$. \rem{more detail}
Thus, we have
$\Val \ec{y} \interp{ t } = \Val \ec{x} \interp{ \Val \ec{x} } = \Val \ec{x}$, and thus 2 holds for $\Val$.
Statement 3 holds for $\Val$ since \rn{Unique} does not apply.

We now show that $\Val = \nf{(\Val)}$.
Assume by contradition $\Val \ec{x} \neq \nf{(\Val \ec{x})}$ for some $\Val \ec{x}$ of minimal size.
We have that $\Val \ec{x}$ is of the form $\MU {\Varec{x}}.\; \const{C}( t_1, \ldots, t_n )$.
Due to our construction of $\Val$, 
we know $\ec{x}$ contains a constructor $\const{C}( z_1, \ldots, z_n )$ and $\muvar( t_i ) = z_i$ for $i = 1, \ldots, n$.
Since $\Val \ec{x}$ is a minimal, it
contains a subterm of the form $\MU {\Varec{y}}.\; \const{C}( u_1, \ldots, u_n )$
where $\Val \ec{x} \interp{ t_i } \aequiv \Val \ec{x} \interp{ u_i }$ for $i = 1, \ldots, n$.
Due to our construction of $\Val$, $\ec{y}$ contains a constructor $\const{C}( w_1, \ldots, w_n )$ and $\muvar( u_i ) = w_i$ for $i = 1, \ldots, n$.
Since \rn{Cong} does not apply to $\Ec$
we have $\ec{w_j}$ and $\ec{z_j}$ are distinct for some $j$.
By statement 2, $\Val \ec{x} \interp{ t_j } = \Val \ec{z_j}$
and $\Val \ec{x} \interp{ u_j } = \Val \ec{w_j}$,
which are not $\alpha$-equivalent by statement 3, contradicting the fact that $\MU {\Varec{y}}.\; \const{C}( u_1, \ldots, u_n )$ is a bisimilar subterm of $\Val \ec{x}$.
Thus, $\Val = \nf{(\Val)}$, and statements 2 and 3 hold for $\nf{(\Val)}$.

We now show that if statements 2 and 3 hold for some $\Val_1$,
then 2 and 3 also hold for $\nf{(\Val_1 \sigma)}$,
where $\sigma$ is a substitution of the form $\{ {\Varec{x}} \mapsto \mu {\Varec{x}}. t \}$,
${\Varec{x}} \in FV( \Val_1 )$,
and $\mu {\Varec{x}}. t$ is not $\alpha$-equivalent to a term in $\mathcal{T}^\M_{\Varec{x}}(\Val_1)$.
To show 2,
by assumption of 2 on $\Val_1$, we have $\Val_1 \ec{y} \interp{ u } = \Val_1 \ec{x}$ for
all subterms $u$ of $\Val_1 \ec{y}$ where $\muvar(t) = {\Varec{x}}$.
Thus, $\Val_1 \sigma \ec{y} \interp{ u } = \Val_1 \sigma \ec{x}$ and
$\nf{(\Val_1 \sigma)} \ec{y} \interp{ u } = \nf{(\Val_1 \sigma)} \ec{x}$.
To show 3, 
consider two distinct equivalence classes $\ec{y}$ and $\ec{z}$,
and assume by contradiction that $\nf{(\Val_1 \sigma)} \ec{y} \aequiv \nf{(\Val_1 \sigma)} \ec{z}$.
Due to statement 3 for $\Val_1$,
$\ec{y}$ and $\ec{z}$ must have (minimal) subterms 
where $t_1$ occurs in $\Val_1 \ec{y}$ the same position (call it $p$) as $t_2$ occurs in $\Val_1 \ec{z}$,
and $t_1 \not\aequiv t_2$.
%We first show that $\nf{(\Val_1 \sigma)} \ec{y}$ (and for the same reasons, $\nf{(\Val_1 \sigma)} \ec{z}$)
%also contains position $p$.
%If this were not the case,
%then without loss of generality we may assume
%$\Val_1 \sigma \ec{y}$ contains a strict bisimilar subterm 
%$\mu {\Varec{w}}. u$ that strictly contains $t_1 \sigma$.
If $t_1$ (resp. $t_2$) is a free variable that is not ${\Varec{x}}$,
then $\nf{(\Val_1 \sigma)} \ec{y}$ (resp. $\nf{(\Val_1 \sigma)} \ec{z}$)
contains $t_1$ (resp. $t_2$) at position $p$,
and $\nf{(\Val_1 \sigma)} \ec{z}$ (resp. $\nf{(\Val_1 \sigma)} \ec{y}$) does not.
If $t_1$ is of the form $\mu w_1. \const{C}( \vec t )$ 
and $t_2$ is of the form $\mu w_2. \const{D}( \vec u )$,
then the expansion \rem{define} of $\nf{(\Val_1 \sigma)} \ec{y}$ and $\nf{(\Val_1 \sigma)} \ec{z}$
are $\alpha$-disequivalent at position $p$.
Since $t_1$ and $t_2$ are minimal,
say $t_1$ is of the form $\mu w. \const{C}( \vec t )$,
and $t_2$ is ${\Varec{x}}$.
Since $\sigma$ maps ${\Varec{x}}$ to a closed $\mu$-term $\mu {\Varec{x}}. t$,
we have that $FV( t_1 ) \subseteq \{ {\Varec{x}} \}$,
or else the expansion of $\nf{(\Val_1 \sigma)} \ec{y}$ and $\nf{(\Val_1 \sigma)} \ec{z}$
are $\alpha$-disequivalent at position $p$ since they do not contain the same free variables.
Since $\Val_1 \ec{y} \interp{t_1}$ is normal,
there is a closed $\mu$-term $v \in \mathcal{T}^\M_{\Varec{x}}( \Val_1 )$
such that $v \vsimv{\Varec{x}} \Val_1 \ec{y} \interp{t_1}$.
Thus, by assumption on our selection of $\mu {\Varec{x}}. t$,
we have $\mu {\Varec{x}}. t \not\vsimv{\Varec{x}} \Val_1 \ec{y} \interp{t_1}$,
which implies that the expansion of $\nf{(\Val_1 \sigma)} \ec{y}$ and $\nf{(\Val_1 \sigma)} \ec{z}$
are $\alpha$-disequivalent at position $p$.

Thus, by induction on the number of applications of the above rule used to obtain $\Val^\ast$,
we have that $\Val^\ast$ satisfies statements 2 and 3.
\qed
\end{proof}
